---
title: "Technology Innovation with DE optimization"
author: "Celia Marraro"
date:  \`r format(Sys.Date(), "%d, %B, %Y")`\
output: html_document
---
The focus of this project is innovation and technology diffusion within the energy sector.
Specifically, the purpose of the study will be to determine how and why clean-energy 
technology enters into the market, and whether or not it supersedes incumbent technologies. 
The mode of analysis will use evolutionary modeling techniques to describe technology selection 
and the dynamics of selection based on the evolutionary concepts of recombination and mutation. 

The study uses time-series data from 1971 to 2015 from the IEA database, as well as BEA industrial
price indexes for electricity, accessed through the Quandl API, also between 1971-2015:

https://www.iea.org/media/statistics/IEA_HeadlineEnergyData_2017.xlsx

https://www.quandl.com/data/BEA/T50404_A-Price-Indexes-for-Private-Fixed-Investment-in-Structures-by-Type-Annual-data-from-1929-to-2016 _You will need an API key to access the data_


Requirements

```{r, warning=F, message=F}
library(openxlsx)
library(Quandl)
library(ggplot2)
library(lattice)
library(reshape)
library(forecast)
library(zoo)
library(car)
library(plyr)
library(dplyr)
library(RcppDE)
```

Read in Energy Data from IEA 

```{r, warning=F, message=F}
url = "https://www.iea.org/media/statistics/IEA_HeadlineEnergyData_2017.xlsx"

datEnergy <- readWorkbook(url, sheet=4, check.names=T, detectDates=T)
dataUS <- filter(datEnergy, datEnergy[,1]=='United States')
colnames(dataUS) <- datEnergy[1,]
dataUS = dataUS[, -c(1,4:6,52)]
rownames(dataUS) <- 1:NROW(dataUS)
```
Gwh of electricity per technology i.e. Installed capacity

```{r, message=F, fig.width=9, fig.height=9}
years <- 1971:2015
datOutputE <- filter(dataUS, dataUS[,'Flow']=='Electricity output (GWh)') 
                     %>% as.data.frame() %>% reshape( direction="long",
                     varying=list(names(dataUS[3:47])), ids=1:4,
                     times=rep(years), v.names="Energy Output","Year")
datOutputE=datOutputE[,-5]
datCapacity=cast(datOutputE, Year~Product)
plot(datCapacity)
capacity_ts <- xts(datCapacity, 
                   order.by=as.Date.yearqtr(datCapacity$Year))
colnames(capacity_ts) <- c("Fossil Fuel Gwh", "Nuclear Gwh", 
                           "Renewable Gwh", "Total Gwh")
tsRainbow <- rainbow(ncol(capacity_ts))
plot.zoo(capacity_ts[10:NROW(capacity_ts),], col=tsRainbow, 
         main="Installed Capacity by Technology")
capacity_ts <- cumsum(capacity_ts)
```
Import BEA data from Quandl API, price Index & cumulative price index

```{r}
Quandl.api_key(Sys.getenv("Quandl_Auth")) 

priceE <- Quandl("BEA/T50404_A", order="asc", type="xts")
priceE = priceE[,c(1,16:18)]; 
priceE = priceE["1971/2015"]
priceE = xts(priceE, order.by=as.Date.yearqtr(datCapacity$Year))
priceE <- priceE[,3]
colnames(priceE) <-"Industry Electric Price Index"
priceC <- Quandl("BEA/T50404_A", order="asc",
                 type="xts", transform="cumul")
priceC = priceC[,c(1,16:18)]; 
priceC = priceC["1971/2015"]
priceC = xts(priceC, order.by=as.Date.yearqtr(datCapacity$Year))
priceC <- priceC[,3]
colnames(priceC) <-"Cumulative Electric Price Index"
```
Capital Investments (*Primary* inputs in electricity production/technology)

```{r, message=F}
d1 <- filter(dataUS, dataUS[,'Flow']=='Electricity plants (ktoe)') 
             %>% as.data.frame() %>% reshape( direction="long",
             varying=list(names(dataUS[3:47])), ids=1:9,
             times=rep(years), v.names="Input", "Year")
Input_dat <- d1[,-5]
Input_dat <- cast(Input_dat, Year~Product)
Input_dat = Input_dat[,-c(3:5)]
Input_ts <- xts(Input_dat, order.by=as.Date.yearqtr(Input_dat$Year))
Input_ts = Input_ts[,-1]
colnames(Input_ts)=colnames(Input_dat[,2:7])
```
Convert inputs to fractional shares of technology_i, i=1,2,3

```{r, message=F,warning=F}
m <- apply(as.matrix.noquote(Input_ts), 2, as.numeric)
m <- -m
m <- cbind(m, rowSums(m[,1:2]))
m<- m[, c(7,3,5:6)]
shares.dat <- t(m[,c(1:3)])/m[,4]
shares.dat <- t(shares.dat)
shares.dat=as.data.frame(shares.dat)
colnames(shares.dat) <-c("Fossil Fuel Shares", "Nuclear Shares", 
                         "Renewables Shares")
shares.ts <- xts(shares.dat, order.by=as.Date.yearqtr(Input_dat$Year))
shares.ts <- cumsum(shares.ts[,1:3])
```
Emissions data 

```{r, message=F}
datEmissions <- dataUS[nrow(dataUS),]
datEmissions = reshape(datEmissions, direction="long",
                       varying=list(names(datEmissions[3:47])), 
                       ids=1:NROW(datEmissions), times=rep(years), 
                       v.names="Energy Output","Year")
datEmissions= datEmissions[,-5]
datEmissions = cast(datEmissions, Year~Flow)
emission.ts <- xts(datEmissions[,2], order.by=
                   as.Date.yearqtr(datEmissions$Year)) 
                   %>%`colnames<-`("Annual CO2 Emissions")  
```
Main data set and plots

```{r}
datMain.xts <- merge(capacity_ts, shares.ts,
                     priceE, priceC, emission.ts)
xyplot.ts(datMain.xts, col=tsRainbow)
datMain <- as.data.frame.ts(datMain.xts)
scatterplot(datMain$Annual.CO2.Emissions
            ~datMain$Renewables.Shares)
scatterplot(datMain$Annual.CO2.Emissions 
            ~datMain$Fossil.Fuel.Shares)
scatterplotMatrix(datMain[,c(5:7,10)], diagonal="density", 
                  smooth=F, transform=T, smoother=quantregLine, 
                  spread=F)
```
##Autocorrelation tests and forecasting
Renewable Energy

```{r, fig.height=8, fig.width=8}
auto.arima(as.numeric(capacity_ts[,3])) #renewable capacity
fitRenew<- arima(as.numeric(capacity_ts[,3]), order=c(0,2,0))
tsdiag(fitRenew) #time series diagnostic
fCast <- forecast(fitRenew, h=20, level=95)
xyplot(fCast$residuals)
autoplot(fCast)
auto.arima(as.numeric(shares.ts[,3])) #renewable shares
fitShares <- arima(as.numeric(shares.ts[,3]), order=c(0,2,0))
tsdiag(fitShares)
fShCast <- forecast(fitShares, h=20, level=95)
xyplot(fShCast$residuals)
autoplot(fShCast)
```
Emissions

```{r, fig.height=7.7}
plot.ts(emission.ts, type="b")
auto.arima(as.numeric(emission.ts))
fit.emiss <- arima(as.numeric(emission.ts), order=c(0,1,0))
tsdiag(fit.emiss)
fcast.emiss <- forecast(fit.emiss, h=20, level=95)
xyplot(fcast.emiss$residuals, ylab="Emissions", xlab="Years", 
       sub="Emissions' Forecasting Residuals")
autoplot(fcast.emiss)
spec.pgram(as.numeric(datMain.xts[,3]), main="Spectrum of Renewable Energy",
           sub= "Cumulative Installed Capacity (Gwh)")
Acf(coredata(datMain.xts[,c(3)]), na.action=na.pass, lag.max=40, 
    main="Correlogram for Renewable Energy", 
    sub="Cumulative Installed Capacity (Gwh)")
```
##Objective function/Inputs
  + objFun() denotes a unit cost of technology_i at time = t
  + beta, the learning elasticity, is > 0.3 for renewable energy sources and <.03 for incumbents 

```{r}
costDat <- merge(log(datMain.xts[,c(3,2,1,7,6,5)]))
beta<- seq(.022, .4, length.out=100) # "learning elasticity"
k <-1:NROW(costDat)

objFun <- function(x1=coredata(costDat[k,1:3]), x2=coredata(costDat[k,4:6])){ 
  if("Renewables.Shares" %in% names(x2)){
  epsilon = sample(beta[beta > .30],1)*x2  
} else {
  epsilon = sample(beta[beta < .03],1)*x2  
}
  cost1 <- 1/x1 + epsilon
  for (l in 1:3){
    ifelse(cost1[k,l] > log(priceE), Inf, cost1)   
  } 
  return(exp(cost1))
k=k+1
}

xyplot.ts(objFun(), superpose=T, lwd=3)
```
#Construct lower and upper bounds for minimization parameters_
```{r, message=F, warning=F}
set.seed(384)
a1 <- seq(from=.01, to=1.1, length.out=10)
a2 <- seq(from=.02, to=1.5, length.out=10)
a3 <- seq(from=.05, to=4.99, length.out=10)

b1 <- seq(from=.5, to=10.5, length.out=10)
b2 <- seq(from=1, to=11, length.out=10)
b3 <- seq(from=2, to=15, length.out=10)

low <- list(a1,a2,a3)
for(i in 1:length(low)){
  lowBound <-sample(low[[i]], size=1) 
  lowBound <-rep(lowBound,3)
  return(lowBound)
}

upper <- list(b1,b2,b3)
for(j in 1:length(upper)){
  upperBound <- sample(upper[[j]], size=1)
  upperBound <- rep(upperBound, 3)
  if(upperBound>=lowBound) {
}
  return(upperBound)
}
print(lowBound)
print(upperBound)
```

DE optimization

```{r}
set.seed(1235)
mutateZero <- DEoptim(fn=objFun, lower=lowBound, 
                      upper = upperBound, control
                      =DEoptim.control(trace=F, F=0, 
                      CR=.55, storepopfrom=1, 
                      storepopfreq=2, strategy=3, 
                      itermax=500))
summary(mutateZero)

crZero <- DEoptim(fn=objFun, lower=lowBound,
                  upper=upperBound,control
                  =DEoptim.control(trace=F, F=1.5,
                  CR=0, storepopfrom=1, storepopfreq=2,
                  strategy=3, itermax=500))
summary(crZero)

midRates <- DEoptim(fn=objFun, lower=lowBound,
                    upper=upperBound, control 
                    =DEoptim.control(trace=F, F=1,
                    CR=.5, storepopfrom=1, storepopfreq=2,
                    strategy=3, itermax=500))
summary(midRates)

crGreater <- DEoptim(fn=objFun, lower=lowBound,
                     upper=upperBound, control 
                     =DEoptim.control(trace=F, F=.1,
                     CR=.75, storepopfrom=1, storepopfreq=2,
                     strategy=3, itermax=500))
summary(crGreater)

fGreater <- DEoptim(fn=objFun, lower=lowBound,
                    upper=upperBound, control 
                    =DEoptim.control(trace=F, F=1.3,
                    CR=0.08, storepopfrom=1, storepopfreq=2,
                    strategy=3, itermax=500))
summary(fGreater)

zeroRates <- DEoptim(fn=objFun, lower=lowBound,
                     upper=upperBound, control 
                     =DEoptim.control(trace=F, F=0,
                     CR=0, storepopfrom=1, storepopfreq=2,
                     strategy=3, itermax=500))
summary(zeroRates)
```
Summary of convergence results 

```{r}
c(
cat("best:", mutateZero$optim$bestmem, "f:", mutateZero$optim$bestval,"\n"),
cat("best:", crZero$optim$bestmem, "f:",  crZero$optim$bestval,"\n"),
cat("best:", fGreater$optim$bestmem, "f:", fGreater$optim$bestval,"\n"),
cat("best:", midRates$optim$bestmem, "f:", midRates$optim$bestval,"\n"),
cat("best:", crGreater$optim$bestmem, "f:", crGreater$optim$bestval,"\n"),
cat("best:", zeroRates$optim$bestmem, "f:", zeroRates$optim$bestval,"\n"))
```
```{r, fig.width=8.7, fig.height=9}
par(mfrow=c(3,2))
plot(mutateZero, plot.type="bestvalit", log="x",
     type="l", lwd=2, col="blue",
     sub="Strategy: Mutatation(0), Cross Over(0.55)")
plot(crZero, plot.type="bestvalit", log="x",
     type="l", lwd=2, col="darkred",
     sub="Strategy: Cross Over(0), Mutation(1.5)")
plot(midRates, plot.type="bestvalit", log="x",
     type="l", lwd=2, col="green",
    sub="Strategy: Mutation(1) & Cross Over(0.5), Mid-range")
plot(crGreater, plot.type="bestvalit", log="x",
     type="l", lwd=2, col="pink",
     sub="Strategy: Cross Over(0.75) > Mutation(0.1)")
plot(fGreater, plot.type="bestvalit", log="x",
     type="l", lwd=2, col="red",
     sub="Strategy: Cross Over(0.08) < Mutation(1.3)")
plot(zeroRates, plot.type="bestvalit", log="x",
     type="l", lwd=2, col="darkgreen",
     sub="Strategy: Mutation = Cross Over =0")

Lab.palette <- colorRampPalette(c("blue", "orange", "red"), space="Lab")
par(mfrow=c(3,2))
smoothScatter(density(crZero$member$pop), colramp=Lab.palette, 
                      xlab="Best Member/Generation", ylab="Density")
smoothScatter(density(midRates$member$pop), colramp = Lab.palette, 
                      xlab="Best Member/Generation", ylab="Density")
smoothScatter(density(crGreater$member$pop), colramp=Lab.palette, 
                      xlab="Best Member/Generation", ylab="Density")
smoothScatter(density(mutateZero$member$pop), colramp=Lab.palette, 
                      xlab="Best Member/Generation", ylab="Density")
smoothScatter(density(fGreater$member$pop), colramp=Lab.palette, 
                      xlab="Best Member/Generation", ylab="Density")
smoothScatter(density(zeroRates$member$pop), colramp=Lab.palette, 
                      xlab="Best Member/Generation", ylab="Density")
```







