---
title: "Credit Score Analysis"
author: "Celia Marraro"
date:  \`r format(Sys.Date(), "%d %B, %Y")`\
output: html_document
---
Requirements

```{r,warning=FALSE, message=FALSE}
library(caret)
library(lattice)
library(ROCR)
library(e1071)
library(party)
```

Data

```{r}
data("GermanCredit")
datGerCred <- GermanCredit
```

Variable binning

```{r}
datGerCred$amt.fac <- as.factor(ifelse(datGerCred$Amount<=2500,"0-2500", 
                                ifelse(datGerCred$Amount<=5000, 
									   "2600-5000", "5000+")))
datGerCred$age.fac <- as.factor(ifelse(datGerCred$Age<=30,'0-30', 
                                ifelse(datGerCred$Age<=40,
									   '30-40', '40+')))
datGerCred$Duration <- as.factor(ifelse(datGerCred$Duration<=15,"0-15", 
                                 ifelse(datGerCred$Duration<=30, 
								        "16-30", ifelse(datGerCred$Duration<=50, 
										"31-50", ifelse(datGerCred$Duration<=72, 
										"51-72")))))
datGerCred$default <- as.factor(ifelse(datGerCred$Class=="Bad", "1", "0"))
```
Prediction error conditions for default when bad credit := 1: 

  + false positive, when incorrectly predicting an applicant has good credit 

  + false negative, when incorrectly predicting an applicant has bad credit 
  

plots describing data

```{r}
mosaicplot(default~age.fac, col=T, data=datGerCred)
spineplot(Class~age.fac, data=datGerCred)
xyplot(Amount~Age, data=datGerCred)
xyplot(Amount~Age|datGerCred$default, data=datGerCred)
# remove continuous variables with discrete copies
datBin <- datGerCred[,-c(2,5,10)] 
```
Train/Test split

```{r}
datSamp <- sort(sample(nrow(datBin), nrow(datBin)*0.7))
train <- datBin[datSamp,]
test <- datBin[-datSamp,]
```
Full logit to determine variable selection

```{r, warning=F,message=F}
m.log<-glm(train$default~.,data=train,family=binomial)
test$score <- predict(m.log, test, type="response")
pred.log<-prediction(test$score,test$default)
perf.log <- performance(pred.log, "tpr", "fpr")
plot(perf.log, col="light green")
abline(0,1, lty=8, col="red")  # 45 degree line in graph
auc1 <- performance(pred.log, "auc") # calculates area under roc curve
plot(performance(pred.log, "acc")) #cutoff 

#significant variables from previous logit 
m.log_redux <- glm(default~Duration+  
                   InstallmentRatePercentage+
				   ForeignWorker+ 
				   CheckingAccountStatus.0.to.200+ 
				   CheckingAccountStatus.gt.200+ 
				   CheckingAccountStatus.lt.0+ 
				   CreditHistory.NoCredit.AllPaid+ 
				   CreditHistory.ThisBank.AllPaid+ 
				   CreditHistory.Delay+SavingsAccountBonds.lt.100+
				   SavingsAccountBonds.100.to.500+ 
				   OtherDebtorsGuarantors.None+ 
				   OtherDebtorsGuarantors.CoApplicant, 
				   data=train, family=binomial)
test$score <- predict(m.log_redux, test, type="response")
pred.log2 <-prediction(test$score,test$default)
perf.log2 <- performance(pred.log2, "tpr", "fpr")
plot(perf.log2, col="pink", add=T)
abline(0, 1, lty=8, col="red")
auc2 <- performance(pred.log2, "auc") 
plot(performance(pred.log2, "acc")) 
```
##Area under the curve is `r sprintf("%.7f", auc2@y.values) `

Confusion matrix and stats

```{r}
cmat.logit <- xtabs(~round(test$score)+test$default)
cmat.logit <- confusionMatrix(cmat.logit, dnn=c("Prediction", "Label"))
print(cmat.logit)
```
Decision tree model

```{r}
m.tree <-ctree(default~ Duration+InstallmentRatePercentage
               +ForeignWorker+CheckingAccountStatus.0.to.200
			   +CheckingAccountStatus.gt.200
			   +CheckingAccountStatus.lt.0
			   +CreditHistory.NoCredit.AllPaid
			   +CreditHistory.ThisBank.AllPaid+CreditHistory.Delay
			   +SavingsAccountBonds.lt.100
			   +SavingsAccountBonds.100.to.500
			   +OtherDebtorsGuarantors.None
			   +OtherDebtorsGuarantors.CoApplicant, data=train)
plot(m.tree)
result.dat <- as.data.frame(do.call("rbind", 
                            treeresponse(m.tree, 
							newdata=test))) 
test$tscore<-result.dat[,2] 
pred.tree<-prediction(test$tscore,test$default) 
perf.tree <- performance(pred.tree, "tpr", "fpr")
```
Combined plot of logit model performance and tree model performance

```{r}
plot(perf.log2, col='red', main='Logit vs Ctree')
plot(perf.tree, col='dark green', add=TRUE)
abline(0, 1, lty=8, col="grey")
auc3<- performance(pred.tree, "auc") 
plot(performance(pred.tree, "acc")) 
```
#Area under the curve is `r sprintf("%.7f", auc3@y.values) `

Confusion matrix and stats 

```{r}
cmat.tree <-xtabs(~round(test$tscore)+test$default)
cmat.tree = confusionMatrix(cmat.tree) 
print(cmat.tree)

par(mfrow=c(1,2))
fourfoldplot(cmat.tree$table, main="Decision Tree Model")
fourfoldplot(cmat.logit$table, main="Logit Model")

```


