---
title: "Gender Wage Gap "
author: "Celia Marraro"
date: \`r format(Sys.Date(), "%d %B, %Y")`\
output: html_document
 
---

Requirements 

```{r, warning=F, message=F}
library(lattice)
library(ggplot2)
library(car)
library(stargazer, quietly = T)
library(plm)
library(sampleSelection) 
library(sandwich)
library(lmtest)
library(readr)

```
#Source [data](https://github.com/cjmarraro/wageGapData) from github repository


Read in raw data

```{r, message=F, warning=F}
deg.creds_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tagset_edu.csv")

health_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tagset_health.csv")

spouse_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tagset_spouse.csv")

marital_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tagset_marital.csv")

indust_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tagset_indust.csv")

tenure_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tagset_tenure.csv")

occu_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tagset_occu.csv")

ex_occ_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tagset_ex_occu.csv")

occudesire_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tagset_occdesire.csv")

hrs_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tageset_hrs.csv")

wage_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tageset_hrwage.csv")

income_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tagset_income.csv")

age_dat <- read_csv("https://raw.githubusercontent.com/cjmarraro/wageGapData/master/data/csv_dir/tagset_age.csv")

```


process data

```{r}
age_dat <- age_dat[,c("R0000100", "T3966200")]
income_dat = income_dat[,c("R0214700", "R0214800", "R0328000", 
                           "R0482600", "R0782101","R1024001", 
                           "R1410701", "R1778501", "R2141601", "R2350301",
                           "R2722501", "R2971401", "R3279401", "R3559001",
                           "R3897101","R4295101", "R5626201", "R6364601",
                           "R6909701" ,"R7607800", "R8316300", "T0912400", 
                           "T3045300" ,"T3977400")]  
wage_dat = wage_dat[,c("R0047010", "R0263710", "R0446810", 
                       "R0702510" ,"R0945610", "R1256010", 
                       "R1650810", "R1923410","R2318210", "R2526010",
                       "R2925010" ,"R3127800" ,"R3523500", "R3728500",
                       "R4416800", "R5079800" ,"R5165200" ,"R6478000", 
                       "R7005700" ,"R7702900","R8495200" ,"T0986800")]
  
hrs_dat = hrs_dat[,c("R0070500", "R0337800", "R0545600", "R0839900", 
                     "R1087300","R1463000" ,"R1809800" ,"R2171500", 
                     "R2376300", "R2770900", "R3012700", "R3340100", 
                     "R3604400", "R3954600", "R4582900", "R5267100", 
                     "R5912600", "R6578200", "R7192900" ,"R7879600",
                     "T0121900" )]
hrs_dat$HRS.NA <- NA
occu_dat = occu_dat[,c("R0263400", "R0828000", "R0945300", "R1255700",
                       "R1650500","R1923100", "R2317900", "R2525700", 
                       "R2924700", "R3127400", "R3523100" ,"R3728100",
                       "R4193700", "R4587901", "R5270900","R6473700",
                       "R6592900", "R7209600", "R7898000", "T0138400")]
occu_dat$OCC.NA <- NA
occu_dat$OCC.NA1 <- NA
exp_occu_dat = ex_occ_dat[,-1]
occ_desr = occudesire_dat[,-1]
indust_dat = indust_dat[,c("R0046300", "R0263300", "R0446300", "R0702000", 
                           "R0944900","R1255300", "R1650100", "R1922700", 
                           "R2317500" ,"R2525300", "R2924300", "R3127000", 
                           "R3522700", "R3727700", "R4182000", "R4587901", 
                           "R5270500", "R6472100", "R6591300" ,"R7209100", 
                           "R7897500", "T0137900")]
tenure_dat = tenure_dat[,c("R0068710", "R0333221", "R0539410", "R0833810", 
                           "R1081010","R1456710" ,"R1803510", "R2165110", 
                           "R2372510", "R2763410","R3005210", "R3332610",
                           "R3597610", "R3947800", "R4416300","R5079300", 
                           "R5164600" ,"R6477500", "R7005200", "R7702400",
                           "R8494700", "T0986300")]
marital_dat = marital_dat[,c("R0217501", "R0405601", "R0618601", "R0898401",
                             "R1144901", "R1520101", "R1890801", "R2257901", 
                             "R2445301", "R2871000","R3074700", "R3401400",
                             "R3656800", "R4007300", "R4418400", "R5081400", 
                             "R5166700", "R6479300", "R7007000", "R7704300",
                             "R8496700", "T0988500")]
spouse_dat = spouse_dat[,c("R0155500", "R0312710", "R0482910", "R0784300", 
                           "R1024001","R1410701", "R1780701", "R2143801", 
                           "R2352501", "R2724701", "R2973600", "R3281600", 
                           "R3571801", "R3899300", "R4314401","R4996001",
                           "R5650801", "R6374901", "R6917801", "R8325800",
                           "T0920800" )]
spouse_dat$SPO.NA <-NA
health_dat = health_dat[,-1]
deg.creds_dat = deg.creds_dat[,-1]
```
generate column names for time-series

```{r}
mylist = list()
iterColNames <- function(mylist){
    appendYear = c(as.character(79:94),'96','98','00',
                   '02','04','06')
    myColNames = list()  
    for(i in mylist){
          myColNames[[i]] = paste(i, appendYear, sep = ".")
          myColNames = as.list(myColNames)
          myColNames = myColNames[myColNames= as.character(mylist)]
          }
    return(myColNames)
    }
```
Calculate the variables tenure and "total experience" in 2006

```{r}
calc_experience <- (rowSums(tenure_dat))
tot_exp06 <-as.data.frame(calc_experience) 
tenure <- as.data.frame(rowMeans(tenure_dat))
dat_all <- data.frame(age_dat, deg.creds_dat, exp_occu_dat, 
                      tot_exp06,  occ_desr, income_dat,  occu_dat, indust_dat, 
                      tenure_dat, marital_dat, spouse_dat, wage_dat, hrs_dat, 
                      health_dat, tenure)
  dat_all[dat_all == -1] = NA  # Refused 
  dat_all[dat_all == -2] = NA  # Dont know 
  dat_all[dat_all == -3] = NA  # Invalid missing 
  dat_all[dat_all == -4] = NA  # Valid missing 
  dat_all[dat_all == -5] = NA  # Non-interview 
myColNames <-iterColNames(list("INC", "OCC", "IND", 
                               "TEN", "MAR", "SPO", "WAG", "HRS"))
colnames(dat_all) <- c("CaseID","Age", "CREDITS", "EXPT.OCC", 
                       "TOT.EXP06", "DES.OCC", "Race", "Sex", 
                       as.character(as.list(unlist(myColNames))), "PHYS", "WGHT", 
                       "tenure")
```
Subset individuals with bachelors degree or higher

```{r}
dat_ed = (dat_all$CREDITS > 100)
dat_use <-subset(dat_all, dat_ed)
```
compare percentage change of income in 2006 from previous years income

```{r}
years_stagger <- dat_use[,c(9,12,15,20)]
year06 <- dat_use[,30]
wage <- as.matrix((log(year06/years_stagger))/15 )
                   wage[is.infinite(as.numeric(wage))] <- NA
summary(wage)
```
transform to panel data

```{r}
years <- c(1979:1994,1996,1998,2000,2002,2004,2006)
dat_reshape <-reshape(dat_use, direction = "long",
                      varying = list(myColNames$INC, myColNames$OCC,
                      myColNames$IND,myColNames$TEN, myColNames$MAR, 
                      myColNames$SPO, myColNames$WAG, myColNames$HRS),
                      v.names=c("INC", "OCC", "IND", "TEN", "MAR", 
                      "SPO", "WAG", "HRS"), idvar="CaseID", times=rep(years))
```
Additional features

```{r, warning=F, message=F, fig.width=8.5, tidy=TRUE}
dat_panel = dat_reshape
dat_hrs = (dat_panel[, 'HRS'] >=35) # full-time workers only
data_NLSY <- subset(dat_panel, dat_hrs)
attach(data_NLSY)

#occupation dummies
mngmt = as.numeric((OCC >= 1) & (OCC <= 43))
professional = as.numeric((OCC >= 50) & (OCC <= 95))
hard_science = as.numeric((OCC >= 100) & (OCC <= 156))
technical <- as.numeric((OCC >= 203) & (OCC <= 235))
sales <- as.numeric((OCC >= 243) & (OCC <= 285))
clerical <- as.numeric((OCC >= 303) & (OCC <= 395))
service <- as.numeric((OCC >= 403) & (OCC <= 469))
operatives <- as.numeric(( OCC >= 473) & (OCC <= 889))

occ_indx <- as.factor(0*mngmt+1*professional+ 
                      2*hard_science+3*technical+ 
                      4*sales+5*clerical+ 
                      6*service+7*operatives)

levels(occ_indx)[1] <- "mngmt"
levels(occ_indx)[2] <- "prof"
levels(occ_indx)[3] <- "hard science"
levels(occ_indx)[4] <- "technical"
levels(occ_indx)[5] <- "sales"
levels(occ_indx)[6] <- "clerical"
levels(occ_indx)[7] <- "services"
levels(occ_indx)[8] <- "operatives"

occ_indx = relevel(occ_indx, ref="mngmt")
levels(occ_indx)

female <- as.numeric(Sex==2)
hispanic <- as.numeric(Race==1)
AfAm <- as.numeric(Race==2)
race_other <- as.numeric(Race==3)

# Physical abilities proxy
phys <- as.factor(PHYS) 
levels(phys) <- c("Excellent", "Very good", "Good", "Fair", "Poor")
phys = relevel(phys, ref="Good")

# Physical health proxy
hlth <- as.factor(WGHT)
levels(hlth) <- c("overweight", "underweight", "normal", "not ascertained")
hlth = relevel(hlth, ref="normal")

marital <- as.factor(MAR)
levels(marital) <- c("never married", "married", "separated", 
                     "divorced", "remarried", "widowed")
marital = relevel(marital, ref="married")

#5 yr career expectation parameter <-- career expectations in 1979, 
#those who expected to be married (i.e. didn't plan on having a job) := 0, 
#otherwise := 1
exp_career <- as.numeric(EXPT.OCC==c(1,2,4)) 
tenure = round(data_NLSY$tenure)
logHours <- log(HRS)
logHours[is.infinite(logHours)] <- NA

logWage <- log(WAG)
logWage[is.infinite((logWage))] <- NA

# occupational expectations to be compared with actual occupation
desr_mngmt <- as.numeric((DES.OCC >= 1) & (DES.OCC <= 43))
desr_professional <- as.numeric((DES.OCC >= 50) & (DES.OCC <= 95))
desr_hard_science <- as.numeric((DES.OCC >= 100) & (DES.OCC <= 156))
desr_technical <- as.numeric((DES.OCC >= 203) & (DES.OCC <= 235))
desr_sales <- as.numeric((DES.OCC >= 243) & (DES.OCC <= 285))
desr_clerical <- as.numeric((DES.OCC >= 303) & (DES.OCC <= 395))
desr_service <- as.numeric((DES.OCC >= 403) & (DES.OCC <= 469))
desr_operatives <- as.numeric((DES.OCC >= 473) & (DES.OCC <= 889))

desr.occ_indx <- as.factor(0*desr_mngmt + 1*desr_professional + 2*desr_hard_science + 3*desr_technical + 4*desr_sales + 5*desr_clerical + 6*desr_service + 7*desr_operatives)

levels(desr.occ_indx)[1] <- "desr.mngmt"
levels(desr.occ_indx)[2] <- "desr.prof"
levels(desr.occ_indx)[3] <- "desr.hard science"
levels(desr.occ_indx)[4] <- "desr.technical"
levels(desr.occ_indx)[5] <- "desr.sales"
levels(desr.occ_indx)[6] <- "desr.clerical"
levels(desr.occ_indx)[7] <- "desr.services"
levels(desr.occ_indx)[8] <- "desr.operatives"

print(by(data_NLSY$INC, occ_indx, summary)) # income, occupation comparison 
careers <-  as.data.frame(cbind(as.character(occ_indx), 
                                as.character(desr.occ_indx)))
colnames(careers) <- c("Actual", "Desired")
careers_tab <- xtabs(~Actual + Desired, careers, addNA=T)
dActual <- rowSums(careers_tab)
dDesired <- colSums(careers_tab)
career_result <- rbind(dActual, dDesired) 
barplot(career_result[,-9], col=c("black", "grey"), 
        beside=T, main="Actual vs. Desired Occupation", 
        legend.text=c("Actual", "Desired"), ylim=c(0,960))
data_NLSY = cbind.data.frame(data_NLSY[,c("CaseID", "time","Age","Race" ,"Sex", "PHYS",
                                          "WGHT", "CREDITS","EXPT.OCC", "TOT.EXP06", 
                                          "DES.OCC","tenure","INC", "OCC" , "IND", 
                                          "MAR",    "SPO","WAG","HRS")] ,occ_indx , 
                                          desr.occ_ind, marital,female, hispanic,AfAm,
                                          race_other, logHours,logWage, hlth, exp_career)

detach(data_NLSY)
#Tests
coplot(logWage ~ time|CaseID*female, type="l",
       col = c("blue", " pink"), data=data_NLSY,  
       panel=function(x,y,...) panel.smooth(x,y,span=.6,...))
xyplot(logWage~time, data=data_NLSY, 
       groups=factor(female,labels=c('M','F')), 
       pch=20,auto.key=T,type=c("p","g"))
```
Tests

```{r}
m.ols_null <- lm(logWage~occ_indx+female)
m.ols_full <- lm(logWage~female+AfAm+hispanic+
                 phys+hlth + marital+occ_indx+
                 tenure+I(tenure^2)+data_NLSY$TOT.EXP06+
                 I(data_NLSY$TOT.EXP06^2)+exp_career+IND+
                 SPO, data=data_NLSY[,c(1:11,13:19)])
mfixed <- plm(logWage~occ_indx+female, data=data_NLSY,
              index=c("CaseID", "time"), 
              model="within") #F stat < .05, use log wage only
pFtest(mfixed, m.ols_null) # F test p-val < .05; use fixed
mrandom <- plm(logWage~occ_indx+female,
               data=data_NLSY, index=c("CaseID", "time"),
               model="random")
phtest(mfixed, mrandom) # < .05, use fixed
mfixed.2 <-plm(logWage~female+AfAm+hispanic+phys+ 
               hlth+marital+occ_indx+tenure+I(tenure^2)+ 
               TOT.EXP06+I(TOT.EXP06^2)+
               exp_career + IND+SPO, 
               index=c("CaseID", "time"),
               model="within", data=data_NLSY)
coeftest(mfixed.2)
coeftest(mfixed.2, vcovHC(mfixed.2, method = "arellano"))
```
#Selection bias
```{r, warning=F,message=F}
exp_career_female <- exp_career*female
selection.1 <- selection(selection=exp_career_female~occ_indx + marital, 
                         outcome=logWage~tenure+I(tenure^2)+data_NLSY$SPO+ 
                         data_NLSY$TOT.EXP06, data=data_NLSY[,c(1:11,13:19)], 
                         method="2step" )
probit_sel <- glm(exp_career_female~occ_indx+AfAm+hispanic+logHours+ 
                  tenure+I(tenure^2)+data_NLSY$TOT.EXP06+I(data_NLSY$TOT.EXP06^2)+
                  data_NLSY$SPO+marital, family="binomial" (link="probit"), 
                  data=data_NLSY[,c(1:11,13:19)])
imr_var <- dnorm(probit_sel$linear.predictors)/pnorm(probit_sel$linear.predictors)
IMR <-cbind(imr_var, exp_career_female) #inverse mill's ratio
outcome1 <- lm(logWage ~ tenure + I(tenure^2)+
               data_NLSY$SPO+data_NLSY$TOT.EXP06+
               IMR[,1], subset=(exp_career_female==1),
               data=data_NLSY[,c(1:11,13:19)])
outcome2 <- lm(logWage~logHours+AfAm+
               hispanic+phys+hlth+marital+
               occ_indx+tenure+I(tenure^2)+
               data_NLSY$TOT.EXP06+
               I(data_NLSY$TOT.EXP06^2)+
               data_NLSY$SPO+IMR[,1],
               subset=(exp_career_female==1), 
               data=data_NLSY[,c(1:11,13:19)])
```
#Tables
```{r, results='asis', warning=F, message=F}
star_probit <- stargazer(probit_sel, type="html", 
                         out="tableprobit.html", align=T, 
                         title="Table. 1 Probit Model of Occupational Expectations for Women", 
                         intercept.bottom=F, flip=T)
star_selection <- stargazer(m.ols_null, m.ols_full, 
                            outcome1, outcome2, selection.1, 
                            type="html", out="tableselection.html",
                            title="Table. 2 Heckman Model Selection", 
                            column.labels=c("uncorrected", "corrected"), 
                            column.separate=c(2,2), selection.equation=T, 
                            single.row=T, dep.var.labels.include=F, model.names=T, 
                            align=T, multicolumn=F)
star_FE <- stargazer(m.ols_null, m.ols_full, mfixed, mfixed.2, 
                     type="html", out="tableFE.html", 
                     title="Table. 3 OLS and Fixed Effects Comparison",
                     column.labels=c("OLS", "FE"), column.separate=c(2,2), 
                     single.row=T, dep.var.labels.include=F, model.names=F, 
                     align=T, multicolumn=F)
#robustness 
cov1 <- vcovHC(mfixed.2)
robust.se <- sqrt(diag(cov1))
star_Robust <- stargazer(mfixed.2, title="Table. 4 Robust Standard Errors", 
                         type="html", out="robust", single.row=T, 
                         se=list(robust.se, NULL))
coeftest(mfixed.2)
````


